---
title: "Certara MCP Server: Setup and Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Certara MCP Server: Setup and Usage}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Quick start {.tabset .tabset-fade .tabset-pills}

Already have `Certara.R` installed? Configure your assistant with a single call,
then reload the client. These use `scope = "user"` (enabled in every current and
future project) and `run = TRUE` (register via the client CLI when applicable).
Pick your client below. New to the details? Read on past the quick start.

### Cursor

```{r, eval = FALSE}
Certara.R::write_mcp_config(client = "cursor", scope = "user", run = TRUE)
```

### Claude Code

```{r, eval = FALSE}
Certara.R::write_mcp_config(client = "claude-code", scope = "user", run = TRUE)
```

### Claude Desktop

```{r, eval = FALSE}
Certara.R::write_mcp_config(client = "claude-desktop", scope = "user", run = TRUE)
```

### Codex

```{r, eval = FALSE}
Certara.R::write_mcp_config(client = "codex", scope = "user", run = TRUE)
```

## What the Certara MCP server is

`Certara.R` hosts an optional, federated [Model Context Protocol
(MCP)](https://modelcontextprotocol.io/) server that exposes the Certara
ecosystem's knowledge base and tool providers -- NLME modeling via
`Certara.RsNLME`, machine-learning model search via `Certara.RDarwin`, VPC via
`tidyvpc`, and more -- to AI coding assistants.

Supported clients:

* **Cursor**
* **Claude Code** -- the `claude` command-line agent (terminal / CLI)
* **Claude Desktop** -- the desktop app, including its **Code** tab
* **Codex**

The MCP server is **optional**. Building, fitting, and reporting models with
`Certara.RsNLME` and the rest of the suite does not require it; it is a
convenience layer for AI-assisted workflows. When enabled, agents can author and
validate PML, launch and watch fits asynchronously, run diagnostics, and emit
audit-ready R scripts, all grounded in the same knowledge base the packages ship.

## Prerequisites and installation

The MCP stack ships as dependencies of `Certara.R`. Installing the ecosystem
meta-package pulls everything the server needs:

```{r, eval = FALSE}
install.packages(
  "Certara.R",
  dependencies = TRUE,
  repos = c(
    "https://certara.jfrog.io/artifactory/certara-cran-release-public/",
    "https://cloud.r-project.org"
  )
)
```

RsNLME users who install the full suite via the `Certara_Packages` vector in the
[RsNLME Installation
Guide](https://certara.github.io/R-RsNLME/articles/installation.html) already
have `Certara.R`.

The MCP layer has three required helpers with version floors:

* `btw (>= 1.2.0)`
* `mcptools (>= 0.2.0)`
* `ellmer (>= 0.4.0)`

A normal release install pulls these automatically. A developer install with
`repos = NULL` (installing a source tarball directly) skips `Suggests` and can
leave them missing; install them explicitly in that case:

```{r, eval = FALSE}
install.packages(c("btw", "mcptools", "ellmer"))
```

The provider packages that require Certara's NLME Engine (for example
`Certara.RsNLME`) still need the engine installed and `INSTALLDIR` set; see the
[RsNLME Installation
Guide](https://certara.github.io/R-RsNLME/articles/installation.html).

## Configure a client with `write_mcp_config()`

`write_mcp_config()` writes (or merges) the MCP server entry into the client's
own configuration file and, where supported, pre-authorizes the Certara tools.
You run it once per client from an R console; the client then launches the
server automatically.

```{r, eval = FALSE}
# Cursor, project scope -- writes/merges .cursor/mcp.json
Certara.R::write_mcp_config(client = "cursor", scope = "project")

# Claude Code, project scope -- writes/merges .mcp.json
Certara.R::write_mcp_config(client = "claude-code", scope = "project")

# Claude Code, user scope -- registered via the claude CLI when on PATH
Certara.R::write_mcp_config(client = "claude-code", scope = "user", run = TRUE)

# Claude Desktop (user scope) -- merges claude_desktop_config.json
Certara.R::write_mcp_config(client = "claude-desktop")

# Codex -- writes the managed block to ~/.codex/config.toml directly
Certara.R::write_mcp_config(client = "codex")
```

What each client writes:

* **Cursor** -- the server entry in `.cursor/mcp.json` (project) or
  `~/.cursor/mcp.json` (user), plus a `.cursor/rules/certara-mcp-usage.mdc`
  routing rule at project scope.
* **Claude Code** -- `.mcp.json` at project scope; user/local scopes are managed
  through the `claude` CLI. A `certara-mcp-usage.md` guidance doc is imported via
  `CLAUDE.md`.
* **Claude Desktop** -- merged into `claude_desktop_config.json` (user scope
  only; on Windows the MSIX-virtualized path Desktop actually reads). Fully quit
  and relaunch Desktop after a config change. On Windows, also see
  [Claude Desktop on Windows: required environment variables](#claude-desktop-on-windows-required-environment-variables)
  -- MCP-launched fits need a few variables added to the `env` block.
* **Codex** -- a managed block in `~/.codex/config.toml` (the only way to set
  Codex per-tool approvals); the equivalent `codex mcp add` command is printed
  for reference.

Key arguments:

* `scope` -- where the configuration is written, which controls where the server
  is available:
    * `"project"` (default) -- writes into the current project folder (e.g.
      `.cursor/mcp.json`), so the server is enabled **only when you open that
      folder**. Choose this for per-analysis setup that can be version-controlled
      with the project.
    * `"user"` -- writes into your home-directory config (e.g.
      `~/.cursor/mcp.json`), so the server is enabled **globally -- in every
      current and future project** on your machine.
    * `"local"` (Claude Code only) -- a private, per-project entry in
      `~/.claude.json` that is not shared with collaborators through the
      project's `.mcp.json`.
* `tool_allowlist` -- when `TRUE` (default), pre-authorize the Certara tools so
  they run without per-tool approval prompts, using each client's native
  mechanism (Cursor `permissions.json`, Claude Code `settings.json`, Codex
  approval settings). Claude Desktop has no config-file allowlist, so approve
  tools in the UI per session.
* `run` -- controls whether the CLI-managed clients are configured *for* you. By
  default (`FALSE`) `write_mcp_config()` only **prints** the client CLI command
  so you can inspect and run it yourself. Setting `run = TRUE` executes that
  command when the client CLI is on your `PATH` (currently the Claude Code
  user/local scopes, via `claude mcp add`), falling back to printing if the CLI
  is not found. File-based targets (Cursor, Claude Code project scope, Codex,
  Claude Desktop) are always written directly, so `run` does not affect them.

Use `remove_mcp_config()` to undo a configuration:

```{r, eval = FALSE}
# Remove the Cursor project-scope server written above.
Certara.R::remove_mcp_config(client = "cursor", scope = "project")
```

## Claude Desktop on Windows: required environment variables

> **Quick tip.** Short on time? Point your Claude session at this article (share
> the page, or paste this section) and ask it to *"diagnose and fill in the
> environment variables that Claude Desktop sanitizes out of my
> `claude_desktop_config.json`."* With filesystem access the agent can run the
> env-check job below, see which variables are empty, and add the `env` block to
> your config for you -- you then just **fully quit and restart Claude Desktop**
> for it to take effect.

**This applies to Claude Desktop on Windows when you launch NLME fits through the
MCP job tools** (`start_nlme_fit_spec`, `start_nlme_job`, `start_nlme_fitmodel`,
the SCM scripts, etc.).

**Symptom.** MCP-launched fits fail at the compile stage with
`Compile/Link Failed`, even though the *same model fits fine* from RStudio or a
plain `Rscript` run in a terminal.

**Cause.** Claude Desktop launches MCP servers with a **sanitized, minimal
environment** -- it does not pass your full user/system environment to the server
process. On Windows a child process inherits a **frozen snapshot** of its
parent's environment (it never re-reads the registry), and the job launcher
copies the server's environment into every spawned fit. So whatever Claude
Desktop omits is missing in the server *and* in each fit child:

```text
Claude Desktop  ->  Rscript.exe (MCP server)  ->  job child (Rscript -> PowerShell -> TDL5/gcc)
```

The critical omission is **`ComSpec`**: the engine's PowerShell wrapper shells
out with `& $env:comspec /c ...`, and with `ComSpec` empty every shell-out fails,
which surfaces as `ERROR in generating Model.cpp` -> `Compile/Link Failed`.
`NLMEGCCDir64` and `INSTALLDIR` are also typically dropped, so the compile either
cannot find `TDL5.exe`/engine libraries or falls back to an incompatible gcc
(e.g. RTools) instead of the Certara-shipped GCC 8.4.0.

**Fix.** Declare the variables explicitly in the server's `env` block in
`claude_desktop_config.json`, then fully restart Claude Desktop:

```json
"mcpServers": {
  "certara-r": {
    "command": "C:\\Program Files\\R\\R-4.6.0\\bin\\x64\\Rscript.exe",
    "args": [
      "-e",
      "Certara.R::launch_certara_mcp(btw_groups = c('docs'), session_tools = FALSE, server_name = 'certara-r', job_watch_wait_seconds = 600, tool_profile = 'full')"
    ],
    "type": "stdio",
    "env": {
      "ComSpec": "C:\\WINDOWS\\system32\\cmd.exe",
      "INSTALLDIR": "C:\\Program Files\\Certara\\NLME_Engine",
      "NLMEGCCDir64": "C:\\Program Files\\Certara\\mingw64\\",
      "TMP": "C:\\Users\\<you>\\AppData\\Local\\Temp",
      "NUMBER_OF_PROCESSORS": "<your core count>",
      "PATHEXT": ".COM;.EXE;.BAT;.CMD;.VBS;.VBE;.JS;.JSE;.WSF;.WSH;.MSC",
      "PhoenixMSMPIDir": "C:\\Program Files\\Certara\\MPI\\"
    }
  }
}
```

Notes:

* The `command` and `args` are the ones `write_mcp_config("claude-desktop")`
  already wrote -- you are only **adding** the `env` block. Edit the exact file
  the call reported (on Windows this may be the MSIX-virtualized path, not the
  one Desktop's "Edit Config" button opens; see the Claude Desktop bullet under
  *Configure a client* above).
* Replace `<you>` and `<your core count>` with your values.
* `ComSpec` is the one that actually unblocks `Compile/Link Failed`; the gcc
  compatibility warning is a red herring, but set `NLMEGCCDir64` too so the
  Certara GCC 8.4.0 is used. `INSTALLDIR` must point at your NLME-Engine install.
* `TMP`, `NUMBER_OF_PROCESSORS`, and `PATHEXT` are commonly dropped as well
  (`TEMP`/`USERPROFILE`/`SystemRoot` usually survive); include them to be safe.
  The full default allowlist Claude Desktop passes on Windows is `APPDATA`,
  `HOMEDRIVE`, `HOMEPATH`, `LOCALAPPDATA`, `PATH`, `PROCESSOR_ARCHITECTURE`,
  `SYSTEMDRIVE`, `SYSTEMROOT`, `TEMP`, `USERNAME`, and `USERPROFILE` -- anything
  else must be declared explicitly.
* **`PhoenixMSMPIDir` (optional -- for MPI parallelization).** Locates the
  MS-MPI runtime that RsNLME execution hosts use when
  `hostParams(parallelMethod = "LOCAL_MPI")` (the directory holding `mpiexec.exe`
  / `msmpi.dll`). Unlike the variables above, **its absence is silent, not an
  error**: the run does not fail -- the job log prints
  `MPI not found on the system. Using localhost without parallelization.` and
  proceeds **serially**. A single small fit barely notices, but SCM
  stepwise/shotgun covariate searches and bootstraps lose all scenario-level
  parallelism, so MCP-launched runs come out much slower than the identical run
  from RStudio/terminal. Add it (value `C:\Program Files\Certara\MPI\`, trailing
  backslash) to restore parallel execution; a correctly-wired run instead logs
  `Using MPI host with N cores` and `[MPI plan] ... method=local_mpi`.
* **Restart is required.** Claude Desktop reads MCP config only at startup --
  fully quit it (system tray -> Quit, not just closing the window) and relaunch.

**Diagnose it yourself.** Run a throwaway MCP job that prints what the child
actually sees, then read `stdout.log` in the returned `run_dir` (empty values
mean the variables are not reaching the fit children):

```{r, eval = FALSE}
start_nlme_job(
  expr = "cat('ComSpec=', Sys.getenv('ComSpec'), '\n');
          cat('INSTALLDIR=', Sys.getenv('INSTALLDIR'), '\n');
          cat('NLMEGCCDir64=', Sys.getenv('NLMEGCCDir64'), '\n');
          cat('PhoenixMSMPIDir=', Sys.getenv('PhoenixMSMPIDir'), '\n')",
  label = "env-check", project_dir = "<your project>"
)
```

`<run_dir>/artifacts/GEN*/Shared/compilelog.txt` shows the actual compile-stage
error rather than the generic `Compile/Link Failed`. For parallelization, the
job's stderr distinguishes the two states directly: `MPI not found on the
system. Using localhost without parallelization.` (serial) versus `Using MPI
host with N cores` / `[MPI plan] ... method=local_mpi` (parallel).

**Claude Code (CLI) generally does not need this.** A terminal-launched MCP
server inherits the shell's full environment, so `ComSpec` and the Certara
toolchain variables are already present. Caveats: a terminal only snapshots the
environment when it opens (open a new one after changing system variables), and
IDE-embedded terminals inherit from the IDE process (restart the IDE). If a
launcher ever sanitizes the environment, the same `env`-block fix applies to
`.mcp.json` / `claude mcp add`.

## Tool profiles

The launch command bakes in a curated tool subset so the agent's tool list stays
focused. Pass `tool_profile` to `write_mcp_config()`:

* **`full`** (default) -- every tool every provider offers.
* **`core`** -- read/author/validate and data inspection only; no job launches.
* **`authoring`** -- `core` plus model comparison; still no job launches.
* **`execution`** -- `core` plus fit/search job launches and their
  results/qualification tools.
* **`diagnostics`** -- `execution` plus comparison and interpretation.

`find_certara_tools()`, `certara_mcp_capabilities()`, and
`certara_session_status()` are present in every profile.

`full` is the default and is a fine choice most of the time -- there is no
correctness penalty to exposing every tool. Narrow the profile when you want to:

* **reduce tool overload** -- a shorter, task-shaped tool list is easier for the
  agent to reason over and lowers the chance it reaches for a wrong or heavier
  tool;
* **gate risky actions** -- `core` and `authoring` expose no job-launch tools
  (`start_*`), so the agent can read, author, and validate models but cannot
  kick off fits or searches on its own; and
* **trim the startup/latency footprint** on constrained clients.

Pick the narrowest profile that still contains the tools your task needs, and
keep `full` when you want the complete surface available.

## `btw_groups` and live-session tools

`btw_groups` controls which general-purpose `btw` R tool groups are exposed
alongside the Certara tools. The default `"docs"` provides read-only R
documentation lookup and is the leanest, lowest-latency option. Other useful
groups include `"pkg"` (package-development actions), and -- for a live R session
bridged with `btw::btw_mcp_session()` -- `"env"` (inspect that session's objects)
and `"run"` (execute R code there via `btw_tool_run_r`).

```{r, eval = FALSE}
# Expose docs + package-dev tools, plus a bridge to a live RStudio session.
Certara.R::write_mcp_config(
  client        = "cursor",
  btw_groups    = c("docs", "pkg", "env", "run"),
  session_tools = TRUE
)
# Then, in the RStudio session you want the agent to reach:
btw::btw_mcp_session()
```

`session_tools = TRUE` exposes the live-session bridge but does not by itself
connect a session -- after writing the config, reload the MCP server, run
`btw::btw_mcp_session()` in the target session, and have the agent attach to it.

Three R contexts are in play and are easy to confuse:

* the **server process** that answers control-plane tool calls,
* the **bridged live session** (`env` inspection and `run` execution), and
* the **per-job child process** launched by the `start_nlme_*` tools.

`certara_session_status()` reports which of these are active and how to attach a
session.

**When to enable it.** Keep `session_tools = FALSE` (the default) for the common
case: the server runs in its own clean R process, starts faster, and cannot
disturb your interactive workspace. Set `session_tools = TRUE` when you want the
agent to work against *your* live RStudio session -- to read objects you have
already loaded (a prepared dataset, a fitted model) or to run R code in that
session -- instead of re-loading everything in the isolated server process. Pair
it with `btw_groups` `"env"` (inspect objects) and/or `"run"` (execute code).

## Your first MCP session

1. **Reload the client** so it picks up the new configuration (in Cursor:
   Settings -> Tools & MCPs, then enable the Certara R server).
2. **Call `certara_mcp_capabilities()` first.** It reports the concurrency
   policy, the Certara-first rule, the discovered tool/KB providers, the active
   tool profile, and the merged workflow phases (the canonical routing map).
3. **Mind Cursor's routing key.** The key in `.cursor/mcp.json` is `certara-r`,
   but Cursor's tool-call API addresses it as `user-certara-r`. The
   project-scope `.cursor/rules/certara-mcp-usage.mdc` records this, and
   `certara_mcp_capabilities()` reports the mapping under `client_routing`.
4. **Look up knowledge guidance-first.** For "what should I do" / regulatory /
   fit-for-purpose questions, call `guide_pharmacometrics(task, intended_use)`
   before reaching for reference lookups (`search_certara_kb()`,
   `lookup_pml_symbol()`) or `btw` documentation tools.

## Approving tool calls

By default, coding assistants prompt you to approve **every** tool call.
Depending on your client and settings, the agent may ask you to "Approve" the
same Certara tools over and over -- which, for a server you trust and use
constantly, quickly becomes a full-time job of clicking *Approve*. The durable
fix is to **allowlist the whole `certara-r` server** so its tools run without
prompts.

`write_mcp_config(tool_allowlist = TRUE)` (the default) already sets this up for
the clients that support a config-file allowlist:

* **Cursor** -- merges `certara-r:*` into `permissions.json`; takes effect once
  you enable Cursor Run Mode (Allowlist or Auto-review).
* **Claude Code** -- merges `mcp__certara-r` into `permissions$allow` of the
  scope-appropriate `settings.json` (`~/.claude/settings.json` for user,
  `<project>/.claude/settings.json` for project,
  `<project>/.claude/settings.local.json` for local).
* **Codex** -- writes per-tool approvals into `~/.codex/config.toml` (most tools
  approved; privileged cleanup and arbitrary-R job tools stay prompt-gated).

To do it by hand for Claude Code, add the server-wide rule to `settings.json`:

```json
{
  "permissions": {
    "allow": ["mcp__certara-r"]
  }
}
```

`mcp__certara-r` authorizes **every** tool from the server. To approve a single
tool instead, use the fully-qualified form `mcp__certara-r__<tool>` (e.g.
`mcp__certara-r__certara_mcp_capabilities`). Note that Claude Code does **not**
support `*` wildcards for MCP tools, so use the bare server rule
`mcp__certara-r` for "all tools" rather than `mcp__certara-r__*`. Reload the
client after editing.

**Claude Desktop is the exception.** It has no working config-file allowlist --
"Always allow" / "Allow for all tasks" does not persist across sessions (upstream
[claude-code#24433](https://github.com/anthropics/claude-code/issues/24433),
[#56954](https://github.com/anthropics/claude-code/issues/56954)), which is why
`write_mcp_config("claude-desktop")` only prints a note instead of writing one.
If the Desktop **Code** tab honors `~/.claude/settings.json`, the `mcp__certara-r`
rule above silences the prompts; if it uses the desktop app's own agent path
instead, you are limited to clicking "Always allow" per session until Anthropic
ships a persistent setting.

## Inventory, status, and removal

Two read-only console helpers distinguish what is *configured* from what is
*running*:

```{r}
# What Certara MCP entries are written to client config files right now?
Certara.R::list_certara_mcp_configs(client = "cursor", project_dir = tempdir())
```

```{r, eval = FALSE}
# What Certara MCP server processes has this machine actually started?
Certara.R::list_certara_mcp_servers()
```

These are ordinary R functions you call from the R console (or RStudio), not MCP
tools. A server that is configured but never listed as running usually means the
client has not been reloaded since `write_mcp_config()`. `certara_session_status()`
is likewise callable from the console, but its fields read `NA` unless it runs
*inside* a live server (the launch options it reports are recorded only at server
startup) -- in practice the agent calls it as a tool to report the live wiring
(server process, bridged session, active profile).

## Troubleshooting

MCP setup friction is mostly an *environment* problem, not something another tool
can fix. Start with `certara_mcp_capabilities()` and `certara_session_status()`
to see what is actually wired, then work through the common cases:

* **MCP tools missing / server not listed** -- `write_mcp_config()` was not run,
  or the client was not reloaded. Re-run it, reload the client, and enable the
  server in the client's tool settings.
* **`btw` / `mcptools` / `ellmer` not found** -- a `repos = NULL` source install
  skipped `Suggests`. Install from the release repo, or
  `install.packages(c("btw", "mcptools", "ellmer"))` (floors: `btw >= 1.2.0`,
  `mcptools >= 0.2.0`, `ellmer >= 0.4.0`).
* **Engine / compile failures** -- `INSTALLDIR` / NLME not resolvable. Confirm
  the NLME Engine install and that `INSTALLDIR` is set (see the RsNLME
  installation guide).
* **Claude Desktop on Windows: `Compile/Link Failed`** even though the same
  model fits from RStudio -- Desktop sanitizes the server environment. Add
  `ComSpec`, `INSTALLDIR`, `NLMEGCCDir64`, `TMP`, `NUMBER_OF_PROCESSORS`, and
  `PATHEXT` to the `env` block; see
  [Claude Desktop on Windows: required environment variables](#claude-desktop-on-windows-required-environment-variables).
  The same case is summarized in the [RsNLME troubleshooting
  guide](https://certara.github.io/R-RsNLME/articles/troubleshooting.html#claude-desktop-on-windows-extra-environment-variables).
* **Agent behavior feels "old"** -- the installed package predates the current
  MCP knowledge base / rules. Install the current release and re-run
  `write_mcp_config(client = "cursor", scope = "project")` so the Cursor rule is
  refreshed too.
* **Too many tools** -- normal when `btw` groups are enabled. Reason by task
  using `certara_mcp_capabilities()$workflows`; for the leanest server use
  `btw_groups = "docs"` or a narrower `tool_profile`.

There is no automated "doctor" tool -- this checklist plus the two status tools
is the supported path.

## Memory (optional)

The Certara MCP tools can use a local, **opt-in** memory of your preferences,
lessons, and past runs, stored on your machine. The memory is *consumed by the
agent* (its `get_user_preferences` / `get_lessons` / `record_lesson` tools read
and write it so behavior carries across sessions), but the functions below are
**for you, the human**, to control whether that happens and to audit what is
kept. They are ordinary R console functions; none are required to use the MCP
server.

Run them under these conditions:

* `enable_memory()` -- once, if you *want* the agent to remember preferences and
  lessons between sessions. Pass `redact =` a vector of regex patterns to scrub
  sensitive identifiers (compound or project names) from stored free text.
  Memory stays off until you call this.
* `disable_memory()` -- to turn it back off (existing records are kept but not
  used).
* `list_memory_records()` -- to see what is stored (`preferences`, `lessons`,
  `run_memory`, `gaps`, and whether memory is `enabled`).
* `export_memory()` -- to copy the records out for review or backup.
* `delete_memory_record()` / `clear_memory()` -- to prune a single record or
  wipe everything.

If you never call `enable_memory()`, the agent simply operates without memory --
a perfectly normal way to run.

## See also

* `?Certara.R::write_mcp_config`, `?Certara.R::launch_certara_mcp`
* `?Certara.R::list_certara_mcp_configs`, `?Certara.R::remove_mcp_config`
* [RsNLME Installation
  Guide](https://certara.github.io/R-RsNLME/articles/installation.html)
