---
title: "Exploratory Data Analysis (EDA)"
author: "James Craig"
date: "2023-08-22"
output: html_document
params:
  interactive: TRUE
---

```{r eda-setup, echo=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(dpi=300, echo = TRUE, warning = FALSE, message = FALSE, out.width = "100%", out.height = "50%") 
interactive <- params$interactive
```

# Overview of the data {.tabset .tabset-fade .tabset-pills}

## Summary

::: {.keymessage}
*  Elimination profile appears to be bi-exponential
*  Body weight and gender appear to influence PK
:::


## Data

```{r data-preview, eval=interactive}
library(DT)
finaldat <- read.csv(file='./data/finaldat.csv')
DT::datatable(finaldat, rownames = FALSE, filter = "top", 
          options = list(pageLength = 5, scrollX = TRUE))
```




<br><br>

## Demographics Table {.tabset .tabset-fade .tabset-pills}

```{r dm-table}

dm_table <- readRDS(file='./tables/dmtable.RDS')
dm_table

```

<br><br>

## Mean Concentration {.tabset .tabset-fade .tabset-pills}

```{r mean-conc-table}

mean_conctable <- readRDS(file='./tables/mean_conctable.RDS')
mean_conctable

```

<br><br>

## Time Concentration by Subject {.tabset .tabset-fade .tabset-pills}

::: {.results}
-   Disposition appears bi-exponential
:::

```{r mean-log, eval = !interactive}

time_conc <- readRDS(file='./plots/time_concentration_by_subject.RDS')
time_conc

```

```{r mean-log-interactive, eval = interactive}

time_conc <- readRDS(file='./plots/time_concentration_by_subject.RDS')
time_conc |>
  plotly::ggplotly()

```

<br><br>

## Facet Dose Group {.tabset .tabset-fade .tabset-pills}

::: {.results}
-   Possible gender difference in PK with observed concentrations higher in females
:::

```{r time-conc-dose-grp, eval = !interactive}

time_conc_dose_grp <- readRDS(file='./plots/time_concentration_facet_dose_group.RDS')
time_conc_dose_grp

```

```{r time-conc-dose-grp-interactive, eval = interactive}

time_conc_dose_grp <- readRDS(file='./plots/time_concentration_facet_dose_group.RDS')
time_conc_dose_grp |> 
  plotly::ggplotly()

```

<br><br>

## Facet WT Linear {.tabset .tabset-fade .tabset-pills}

::: {.results}
-   Possible WT influence on PK with higher concentrations in lower body weight subjects
:::

```{r wt-linear, eval = !interactive}

time_conc_wt <- readRDS(file='./plots/time_concentration_by_subject_by_wt_cut.RDS')
time_conc_wt

```

```{r wt-linear-interactive, eval = interactive}

time_conc_wt <- readRDS(file='./plots/time_concentration_by_subject_by_wt_cut.RDS')
time_conc_wt |> 
  plotly::ggplotly()

```
